Distributional Trees and Forests

### 2017-06-22 pkg/disttree/inst/useR2017/slides.Rnw
restructuring of tree plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1474 255f4249-2263-4417-b1ea-8498779e27a7

@@

  # disttree
  if(dt.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = 'green4')
    lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
           col = c('black','grey','green4', 'greenyellow'), lty = 1, bty = 'n')
  }

  # distforest
  if(df.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = 'red')
    lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
    lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
           col = c('black','grey','red', 'orange'), lty = 1, bty = 'n')
  }

  # gamlss
  if(g.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = 'blue')
    lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
    lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
           col = c('black','grey','blue', 'lightblue'), lty = 1, bty = 'n')
  }

  # disttree vs distforest vs gamlss
  if(all.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = 'green4')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = 'red')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = 'blue')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
                        TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
           col = c('black','green4','red','blue'), lty = 1, bty = 'n')
  }
}

### plotting function for one data set and sigma only
plot1sigma <- function(learndata, dt=NULL, df=NULL, g=NULL) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

  plot(y = sp$y-sp$true.mu, x = sp$x, type = "p", col="grey", xlab = "x", ylab = "y",  main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
  lines(x = sp$x, y = sp$true.sigma, type = "l", col = 'black')
  lines(x = sp$x, y = -sp$true.sigma, type = "l", col = 'black')
  lines(x = sp$x, y = sp$fitted.sigma.dt, type = "l", col = 'green4')
  lines(x = sp$x, y = -sp$fitted.sigma.dt, type = "l", col = 'green4')
  lines(x = sp$x, y = sp$fitted.sigma.df, type = "l", col = 'red')
  lines(x = sp$x, y = -sp$fitted.sigma.df, type = "l", col = 'red')
  lines(x = sp$x, y = sp$fitted.sigma.g, type = "l", col = 'blue')
  lines(x = sp$x, y = -sp$fitted.sigma.g, type = "l", col = 'blue')
  legend('topleft', c(TeX('$\\sigma$'), TeX('$\\hat{\\sigma}$ disttree'),
                      TeX('$\\hat{\\sigma}$ distforest'), TeX('$\\hat{\\sigma}$ gamlss')), 
         col = c('black','green4','red','blue'), lty = 1)
}

@@

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{0.5cm}
\begin{figure}[!htb]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<1->{
<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = x, y = ytrue, col = "red", lwd = 1.5, main = "")
@
}
\end{center}
\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<2->{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = x, y = ytrue, col = "red", lwd = 1.5, main = "")
@
}
\end{center}
\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\visible<3->{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
lines(x = x, y = ytrue, col = "red", lwd = 1.5)
lines(x = x, y = ytrue+0.1+var, col = "red", lty = 2)
lines(x = x, y = ytrue-0.1-var, col = "red", lty = 2)
@
}
\end{center}
\endminipage

@@

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{0.1cm}
\begin{figure}[!htb]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd = 1.5)
@
\visible<1->{
<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd = 1.5)
@
}
\end{center}

\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
 \begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = yforest, col = "red", lwd = 1.5, main = "")
@
\visible<2->{
<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd = 1.5)
@
}
\end{center}

\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
  \visible<3->{
  \begin{center}
  \begin{tikzpicture}
  \draw[line width=0.01mm, black] (0,0) rectangle (2.6,2.24);
  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
  \end{tikzpicture}
  \end{center}}
\endminipage

@@

  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n21);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n22);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
  \draw[-, line width=1pt] (n22) -- (n23);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
  \draw[-, line width=1pt] (n22) -- (n24);
\end{tikzpicture}
}
\vspace{0.5cm}\\
\code{randomForest}\\
\code{ranger}\\
\code{party(kit)}\\
$\ldots$}
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.68cm}
\visible<3->{
Distributional Trees and Forests\\
\vspace{1.7cm}
\code{disttree}\\
based on \code{partykit}\\
\vspace{1cm}}
\end{center}
\endminipage
\end{figure}
\end{frame}

@@

\textbf{Distributional:} Specify the complete probability distribution
(including location, scale and shape).

@@

\begin{frame}
\frametitle{Prediction}
\vspace{1cm}
{\bf Tree:}\\
\vspace{0.1cm}
\hspace{0.7cm} Estimate $\hat{\theta}$ on the subsample of the learning data which \\
\hspace{0.7cm} ends up in the same terminal node as the new observation.
%\item Evaluate final node for new observation
%\item Return corresponding distribution (parameters)

\vspace{0.5cm}

{\bf Forest:}\\
\vspace{0.1cm}
\hspace{0.7cm} Estimate $\hat{\theta}$ on the whole learning data but weighted by \\
\hspace{0.7cm} the number of trees in which a learning observation ends up\\
\hspace{0.7cm} in the same terminal node as the new observation.
% but weighted by similarities of the learning data and the new observation over all trees
%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
\end{frame}

\begin{frame}
\frametitle{Fitting a Tree}
\vspace*{-0.4cm}
\begin{center} 
DGP:  $Y \sim \mathcal{N}(\mu(X), \sigma(X))$
<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("","","")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "True parameters:",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "True parameters:",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "True parameters:",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)
@

\setkeys{Gin}{width=0.75\linewidth}
<<plottree_dgp, fig=TRUE, echo=FALSE>>=
pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
plot(fig,
     tp_args = list(FUN = identity, width = 18.7, fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)], pval = FALSE),
     drop_terminal = TRUE, tnex = 1.5)
@
\end{center}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-0.4cm}
Model: \code{disttree(y ~ x)}\\
\setkeys{Gin}{width=0.75\linewidth}
<<plottree_estpar, fig=TRUE, echo=FALSE>>=
pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
plot(tree, drop = TRUE, tnex = 1.5, #FUN = FUN,
     tp_args = list(fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]))
@
\end{center}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-0.4cm}
Model: \code{disttree(y~x)}\\
\setkeys{Gin}{width=0.75\linewidth}
<<plottree_box, fig=TRUE, echo=FALSE>>=
plot(as.constparty(tree), tnex = 1.5,
     tp_args = list(fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]))
@
\end{center}
\end{frame}

@@

    #add arrows below
    arrows(0.45,0.05,0.3,0.05,lwd=1) ## add arrow
    arrows(0.6,0.05,0.75,0.05,lwd=1) ## add arrow

@@

  #add arrows below
  arrows(0.45,0.05,0.25,0.05,lwd=1) ## add arrow
  arrows(0.6,0.05,0.8,0.05,lwd=1) ## add arrow

>>>>

coltree <- hcl(c(128,100,30))
colforest <- hcl(c(10,100,30))
colgamlss <- hcl(c(260,100,30))

@@

  # disttree
  if(dt.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = coltree)
    lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
           col = c('black','grey',coltree, 'greenyellow'), lty = 1, bty = 'n')
  }

  # distforest
  if(df.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = colforest)
    lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
    lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
           col = c('black','grey',colforest, 'orange'), lty = 1, bty = 'n')
  }

  # gamlss
  if(g.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = colgamlss)
    lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
    lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
                        TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
           col = c('black','grey',colgamlss, 'lightblue'), lty = 1, bty = 'n')
  }

  # disttree vs distforest vs gamlss
  if(all.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree vs distforest vs gamlss", xaxt='n', yaxt='n', xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = coltree)
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = colforest)
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = colgamlss)
    legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
                        TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
           col = c('black',coltree,colforest,colgamlss), lty = 1, bty = 'n')
  }
}

@@

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{-0.41cm}
\begin{figure}[!htb]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GLM, echo=FALSE, results=hide>>=
nobs <- 200
## GLM
set.seed(7)
x <- c(1:nobs)/nobs
ytrue <- 1+x*1.5
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<1->{
<<plot_motivation_GLM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
box(lwd=5)
lines(x = x, y = ytrue, col = "red", lwd = 5, main = "")
@
}
\end{center}
\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAM, echo=FALSE, results=hide>>=
## GAM
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3 
y <- ytrue + rnorm(nobs,0,0.3)
@
\visible<2->{
<<plot_motivation_GAM, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
box(lwd=5)
lines(x = x, y = ytrue, col = "red", lwd = 5, main = "")
@
}
\end{center}
\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
\begin{center}
<<motivation_GAMLSS, echo=FALSE, results=hide>>=
## GAMLSS
set.seed(7)
x <- c(1:nobs)/nobs
x <- 2*(x-0.5)
ytrue <- x^3
var <- exp(-(2*x)^2)/2
y <- ytrue + rnorm(nobs, 0, 0.1 + var)
@
\visible<3->{
<<plot_motivation_GAMLSS, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(y=y , x=x, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
box(lwd=5)
lines(x = x, y = ytrue, col = "red", lwd=5)
lines(x = x, y = ytrue+0.1+var, col = "red", lty = 2)
lines(x = x, y = ytrue-0.1-var, col = "red", lty = 2)
@
}
\end{center}
\endminipage

@@

\begin{frame}[fragile]
\frametitle{Motivation}
\vspace{-0.2cm}
\begin{figure}[!htb]
\minipage{0.29\textwidth}
\begin{center}
<<motivation_regtree, echo=FALSE, results=hide>>=
## Reg. Tree
set.seed(7)
kappa <- 12
x <- c(1:nobs)/nobs
ytrue <- ytree <- yforest <- numeric(length = length(x))
for(i in 1:nobs) ytrue[i] <- if(x[i]<1/3) 0.5 else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))
y <- ytrue + rnorm(nobs,0,0.3)
for(i in 1:nobs) ytree[i] <- if(x[i]<1/3) 0.5 else {if(x[i]<2/3) 2 else 1}
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd = 1.5, main = "")
lines(x = x, y = ytree, col = "red", lwd=5)
@
\visible<1->{
<<plot_motivation_regtree, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = ytree, col = "red", lwd=5)
@
}
\end{center}

\endminipage
\visible<2->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
 \begin{center}
<<motivation_randforest, echo=FALSE, results=hide>>=
## Random Forest
for(i in 1:nobs) yforest[i] <- if(x[i]<0.27) 0.5 else { if(x[i]<0.39) 0.5 + 1.5*(plogis((x[i]-0.33)/6*700)) else 1+(1-plogis(kappa*(2*(x[i]-0.2)-1)))}
@
\visible<2->{
<<plot_motivation_randforest, fig=TRUE, echo=FALSE>>=
par(mar=c(2,0,2,0))
plot(x = x, y = y, xaxt='n', yaxt='n', ann=FALSE, col = "darkgrey")
box(lwd=5)
#lines(x = x, y = ytrue, col = "grey", lwd=5, main = "")
lines(x = x, y = yforest, col = "red", lwd=5, main = "")
@
}
\end{center}

\endminipage
\visible<3->{{\LARGE$\rightarrow$}}
\minipage{0.29\textwidth}
  \visible<3->{
  \begin{center}
  \begin{tikzpicture}
  \draw[line width=0.2mm, black] (0,0) rectangle (2.6,2.24);
  \node[rectangle, align=center] (sq) at (1.3, 1.12) {\Huge{?}};
  \end{tikzpicture}
  \end{center}}
\endminipage

@@

  \node[ellipse, fill=HighlightBlue!70, align=center] (n20) at (5, 2) {};
  \node[rectangle, fill=HighlightOrange!70, align=center] (n21) at (4.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n21);
  \node[ellipse, fill=HighlightBlue!70, align=center] (n22) at (5.5, 1) {};
  \draw[-, line width=1pt] (n20) -- (n22);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n23) at (5, 0) {};
  \draw[-, line width=1pt] (n22) -- (n23);
  \node[rectangle, fill=HighlightOrange!70, align=center] (n24) at (6, 0) {};
  \draw[-, line width=1pt] (n22) -- (n24);
\end{tikzpicture}
}
\vspace{0.5cm}\\
\code{randomForest}\\
\code{ranger}\\
\code{party(kit)}\\
$\ldots$}
\end{center}
\endminipage
\hspace{0.65cm}
\minipage{0.28\textwidth}
\begin{center}
\vspace{0.68cm}
\visible<3->{
Distributional trees and forests\\
\vspace{1.7cm}
\code{disttree}\\
based on \code{partykit}\\
\vspace{1cm}}
\end{center}
\endminipage
\end{figure}
\end{frame}

@@

\textbf{Distributional:} \\
\hspace{0.3cm} Specify the complete probability distribution\\
\hspace{0.3cm} (including location, scale and shape).

@@

\begin{frame}
\frametitle{Prediction}
%\vspace{1cm}
{\bf Tree:}\\
\begin{enumerate}
\item Select the subsample of the learning data which ends up in the same terminal node as the new observation.
\item Estimate $\hat{\theta}$ on this subsample.
\end{enumerate}
\vspace{0.5cm}

{\bf Forest:}\\
\begin{enumerate}
\item For each learning observation calculate a weight by counting the number of trees in which it ends up in the same terminal node as the new observation.
\item Estimate $\hat{\theta}$ on the wheighted learning data.
\end{enumerate}
% but weighted by similarities of the learning data and the new observation over all trees
%\item Evaluate final node for new observation in each of the trees and investigate similarities with each of the learning observations
%\item Fit a distributional model for the new observation with different influences of the learning observations defined by weights
\end{frame}

\begin{frame}
\frametitle{Fitting a Tree}
\vspace*{-0.5cm}
\begin{center} 
DGP:  $Y \sim \mathcal{N}(\mu(X), \sigma(X))$
\vspace*{-0.21cm}
<<dgp_tree, echo=FALSE, results=hide>>=
data <- data.frame(x = numeric(0), x = numeric(0), x = numeric(0))
names(data) <- c("x","x","x")
fig <- party(
  partynode(1L,
            split = partysplit(2L, breaks = 0.4),
            kids = list(
              partynode(2L, info = c(
                "n = 200",
                "   True parameters:   ",
                expression(mu == '4'),
                expression(sigma == '1')
              )),
              partynode(3L,
                        split = partysplit(3L, breaks = 0.8),
                        kids = list(
                          partynode(4L, info = c(
                            "n = 200",
                            "   True parameters:   ",
                            expression(mu == '12'),
                            expression(sigma == '3')
                          )),
                          partynode(5L, info = c(
                            "n = 100",
                            "   True parameters:   ",
                            expression(mu == '4'),
                            expression(sigma == '3')
                          )))))),
  data
)

node_inner_ext <- function (obj, id = TRUE, pval = TRUE, abbreviate = FALSE, fill = "white", 
    gp = gpar()) 
{
    meta <- obj$data
    nam <- names(obj)
    extract_label <- function(node) {
        if (is.terminal(node)) 
            return(rep.int("", 2L))
        varlab <- character_split(split_node(node), meta)$name
        if (abbreviate > 0L) 
            varlab <- abbreviate(varlab, as.integer(abbreviate))
        if (pval) {
            nullna <- function(x) is.null(x) || is.na(x)
            pval <- suppressWarnings(try(!nullna(info_node(node)$p.value), 
                silent = TRUE))
            pval <- if (inherits(pval, "try-error")) 
                FALSE
            else pval
        }
        if (pval) {
            pvalue <- node$info$p.value
            plab <- ifelse(pvalue < 10^(-3L), paste("p <", 10^(-3L)), 
                paste("p =", round(pvalue, digits = 3L)))
        }
        else {
            plab <- ""
        }
        return(c(varlab, plab))
    }
    maxstr <- function(node) {
        lab <- extract_label(node)
        klab <- if (is.terminal(node)) 
            ""
        else unlist(lapply(kids_node(node), maxstr))
        lab <- c(lab, klab)
        lab <- unlist(lapply(lab, function(x) strsplit(x, "\n")))
        lab <- lab[which.max(nchar(lab))]
        if (length(lab) < 1L) 
            lab <- ""
        return(lab)
    }
    nstr <- maxstr(node_party(obj))
    if (nchar(nstr) < 6) 
        nstr <- "aAAAAa"
    rval <- function(node) {
        node_vp <- viewport(x = unit(0.5, "npc"), y = unit(0.5, 
            "npc"), width = unit(1, "strwidth", nstr) * 1.3, 
            height = unit(3, "lines"), name = paste("node_inner", 
                id_node(node), sep = ""), gp = gp)
        pushViewport(node_vp)
        xell <- c(seq(0, 0.2, by = 0.01), seq(0.2, 0.8, by = 0.05), 
            seq(0.8, 1, by = 0.01))
        yell <- sqrt(xell * (1 - xell))
        xell <- xell*1.11 - 0.055             # to adapt size of the ellipse to the size with p-value
        lab <- extract_label(node)
        fill <- rep(fill, length.out = 2L)
        grid.polygon(x = unit(c(xell, rev(xell)), "npc"), y = unit(c(yell, 
            -yell) + 0.5, "npc"), gp = gpar(fill = fill[1]))
        grid.text(lab[1L], y = unit(1.5 + 0.5,                  # to adapt position of x to its position with p-value
            "lines"))
        #grid.text(lab[1L], y = unit(1.5 + 0.5 * (lab[2L] != ""), 
        #    "lines"))
        grid.text(lab[2L], y = unit(1, "lines"))
        if (id) {
            nodeIDvp <- viewport(x = unit(0.5, "npc"), y = unit(1, 
                "npc"), width = max(unit(1, "lines"), unit(1.3, 
                "strwidth", nam[id_node(node)])), height = max(unit(1, 
                "lines"), unit(1.3, "strheight", nam[id_node(node)])))
            pushViewport(nodeIDvp)
            grid.rect(gp = gpar(fill = fill[2]))
            grid.text(nam[id_node(node)])
            popViewport()
        }
        upViewport()
    }
    return(rval)
}

class(node_inner_ext) <- "grapcon_generator"

\setkeys{Gin}{width=0.74\linewidth}
<<plottree_dgp, fig=TRUE, echo=FALSE>>=
pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
plot(fig, inner_panel = node_inner_ext,
     tp_args = list(FUN = identity, width = 18, fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1.4)
@
\end{center}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-0.5cm}
Model: \code{disttree(y ~ x)}\\
\vspace*{-0.2cm}
\setkeys{Gin}{width=0.74\linewidth}
<<plottree_estpar, fig=TRUE, echo=FALSE>>=
pal <- rgb(c(0.97, 0.64, 1), c(0.70, 0.83, 1), c(0.30, 0.99, 1))
plot(tree, drop = TRUE, tnex = 1.4, #FUN = FUN,
     tp_args = list(fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]))

@
\end{center}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Fitting a Tree}
\begin{center}
\vspace*{-0.5cm}
Model: \code{disttree(y ~ x)}\\
\vspace*{-0.2cm}
\setkeys{Gin}{width=0.74\linewidth}
<<plottree_box, fig=TRUE, echo=FALSE>>=
plot(as.constparty(tree), tnex = 1.4,
     tp_args = list(fill = pal[c(1, 3)]), 
     ip_args = list(fill = pal[c(2, 3)]))
@
\end{center}
\end{frame}

@@

    #add arrows below
    arrows(0.45,0.09,0.3,0.09,lwd=1) ## add arrow
    arrows(0.6,0.09,0.75,0.09,lwd=1) ## add arrow

@@

  #add arrows below
  arrows(0.45,0.09,0.3,0.09,lwd=1) ## add arrow
  arrows(0.6,0.09,0.75,0.09,lwd=1) ## add arrow
